Analyzing Toronto's Dining Scene: Public Food Safety and Economic Indicators in Focus

### 2024-01-25 Outputs_Folder/Paper/toronto_dinesafe_paper.qmd
Fixed typos and updated figure 3

---
title: "Analyzing Toronto's Dining Scene: Public Food Safety and Economic Indicators in Focus"
author: 
  - Tianen (Evan) Hao
thanks: "Code and data are available at: https://github.com/ShadyEvan4830/Toronto_Dinesafe_DataAnalysis.git"
date: January 21, 2024
date-format: long
abstract: "Nowadays, as the market economy continues to thrive, people have more diverse choices of edible products and are also more aware of their potential health risks. This report uses the open-source statistical programming language R to analyze a data set of food safety inspections at Toronto dining establishments and further uses additional data sets to correlate food safety trends with an economic indicator such as the Consumer Price Index (CPI). The analysis shows a limited reduction in sanitation infractions; this potentially means food safety compliance is improving in the City of Toronto. Additionally, the analysis suggests some potential links between economic downturns and increases in health infractions."
format: pdf
toc: true
number-sections: true
bibliography: bibliography.bib
---

@@

This report conducts an in-depth analysis of Dinesafe data provided by the City of Toronto's open data website OpenDataToronto [@opendatatoronto]. The author uses the open-source statistical programming language R [@citeR] to display trends in the data, focusing on key indicators such as inspection pass rate and changes in infraction details. In addition, this report further examines how economic changes indicated by external factors such as the Consumer Price Index (CPI) are related to changes in public health standards within the food industry. Through analysis, results found that restaurants are more likely to violate food safety regulations, and there is a correlation between CPI fluctuations and infraction behaviours. The trend shows that the number of infraction behaviours will increase when CPI decreases.

@@

## Dinesafe
The Dinesafe dataset is publicly available from the City of Toronto [@opendatatoronto] and has many related variables. In this report, some key variables are mainly selected, including inspection dates, establishment type, establishment status, severity, action, and minimum inspections per year, as @tbl-Sample_of_DineSafe_Data shown. Inspection dates represent the time range for data set inspection; this report mainly selects the range from January 2022 to January 2024 for spot inspection analysis. Establishment type represents the relevant business type when Toronto Public Health inspects, such as a restaurant, food store or food takeout. Establishment status mainly reflects whether an enterprise has passed the food safety inspection. In this data set, only relevant enterprises that successfully passed or passed conditionally are displayed. Although it does not directly reflect the situation of failed companies, the severity variable in the data set shows that these companies that passed also have infraction problems in different situations, and the action variable records the different actions taken by these companies after infractions. Finally, minimum inspections per year serve as a reference for analyzing food safety status and corporate behaviour. Considering the hugeness of the dataset data, please refer to \ref{app:clean-dinesafe-dataset}.
```{r}
#| echo: false
#| label: tbl-Sample_of_DineSafe_Data
#| tbl-cap: Sample of DineSafe Data
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)

@@

```
## Toronto's Dashboard Key Indicators
In addition, this report hopes to find the connection between public health trends and society from a broader perspective. To this end, the report further searched for Toronto's Dashboard Key Indicators dataset through @opendatatoronto. This dataset contains many economic-related indicators that the Toronto government is open to the public; the author selected the relevant period between January 2022 and January 2023 of the Consumer Price Index (CPI), including numerical values and years that are consistent with Dinesafe for analysis. Considering the hugeness of the dataset data, please refer to \ref{app:clean-cpi-dataset}.

@@

## Toronto Dine Safe Pass Status
The report extracted the complete data through R and classified the number of institutions by Establishment Status in the Dinesafe dataset [@TorontoDineSafe2021], specifically showing two categories: "Conditional Pass" and "Pass" as @fig-Number_of_Establishments_by_Status shown. Taken together, the count for the immediate "Pass" category is significantly higher with number of 76436, indicated by the higher bar on the right, indicating that more businesses are fully compliant with food safety inspection standards than those 369 businesses receiving a "Conditional Pass." "Conditional pass" indicates that certain businesses must address specific issues to fully comply with health standards. This visual demonstrates that most businesses inspected adhere to the required health and safety guidelines.

@@

Significantly fewer infractions were labelled major than minor ones, which may indicate that more serious health risks are less likely to occur. Nonetheless, the presence of these infractions indicates the need for improvements to ensure public health.

@@

```
\break
## Actions from Infraction Companies
To further evaluate the actions these violating companies took after being flagged, the report used the "Minor" level with the most violations for analysis and created a bar chart to more clearly visualize the data [@fig-distribution_minor_severity_infractions]. Please note that the coding of this chart refers to the "Top 50 Ggplot2 Visualization" [@rstats_top50ggplot2] website. The chart is dominated by the green proportion, representing the behaviour of "notice to comply." This suggests that in most cases of minor level, immediate closure or fines are not the first action Toronto Public Health considers; instead, businesses are allowed to make amends within a certain period.

>>>>

---
title: "Analyzing Toronto's Dining Scene: Public Food Safety and Economic Indicators in Focus"
author: 
  - Tianen (Evan) Hao
thanks: "Code and data are available at: https://github.com/ShadyEvan4830/Toronto_Dinesafe_DataAnalysis.git"
date: January 21, 2024
date-format: long
abstract: "Nowadays, as the market economy continues to thrive, people have more diverse choices of edible products and are also more aware of their potential health risks. This report analyzes a data set of food safety inspections at Toronto dining establishments and further uses additional data sets to correlate food safety trends with an economic indicator such as the Consumer Price Index (CPI). The analysis shows a limited reduction in sanitation infractions; this potentially means food safety compliance is improving in the City of Toronto. Additionally, the analysis suggests some potential links between economic downturns and increases in health infractions; this initially establishes a correlation between economic contraction and increases in food safety infractions, with the aim of further elucidating these relationships and their implications for public policy."
format: pdf
toc: true
number-sections: true
bibliography: bibliography.bib
---

@@

This report conducts an analysis of Dinesafe data provided by the City of Toronto's open data website OpenDataToronto [@opendatatoronto]. The author uses the open-source statistical programming language R [@citeR] to display trends in the data, focusing on key indicators such as inspection pass rate and changes in infraction details. In addition, this report further examines how economic changes indicated by external factors such as the Consumer Price Index (CPI) are related to changes in public health standards within the food industry. Through analysis, results found that restaurants are more likely to violate food safety regulations, and there is a possible correlation between CPI fluctuations and infraction behaviours. The trend shows that the number of infraction behaviours will increase when CPI decreases.

@@

## Dinesafe
The Dinesafe dataset is publicly available from the City of Toronto [@opendatatoronto] and has many related variables. In this report, some key variables are mainly selected, including inspection dates, establishment type, establishment status, severity, action, and minimum inspections per year, as @tbl-Sample_of_DineSafe_Data shown. Inspection dates represent the time range for data set inspection; this report mainly selects the range from January 2022 to January 2024 for spot inspection analysis. Establishment type represents the relevant business type when Toronto Public Health inspects, such as a restaurant, food store or food takeout. Establishment status mainly reflects whether an enterprise has passed the food safety inspection. In this data set, only relevant enterprises that successfully passed or passed conditionally are displayed. Although it does not directly reflect the situation of failed companies, the severity variable in the data set shows that these companies that passed also have infraction problems in different situations, and the action variable records the different actions taken by these companies after infractions. Finally, minimum inspections per year serve as a reference for analyzing food safety status and corporate behaviour. Considering the hugeness of the dataset data, there is a preview of the data is included in Appendix A; please refer to \ref{app:clean-dinesafe-dataset}.
```{r}
#| echo: false
#| label: tbl-Sample_of_DineSafe_Data
#| tbl-cap: Sample of DineSafe Data
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)

@@

```
## Toronto's Dashboard Key Indicators
In addition, this report hopes to find the connection between public health trends and society from a broader perspective. To this end, the report further searched for Toronto's Dashboard Key Indicators dataset through @opendatatoronto. This dataset contains many economic-related indicators that the Toronto government is open to the public; the author selected the relevant period between January 2022 and January 2023 of the Consumer Price Index (CPI), including numerical values and years that are consistent with Dinesafe for analysis. Considering the hugeness of the dataset, a preview of the data is included in Appendix B; please refer to \ref{app:clean-cpi-dataset}.

@@

## Toronto Dine Safe Pass Status
The report extracted the complete data through R and classified the number of institutions by Establishment Status in the Dinesafe dataset [@TorontoDineSafe2021], specifically showing two categories: "Conditional Pass" and "Pass" as @fig-Number_of_Establishments_by_Status shown. Taken together, based on the large sample size, the difference is significant; the count for the immediate "Pass" category is significantly higher with the number of 76436, indicated by the higher bar on the right, indicating that more businesses are fully compliant with food safety inspection standards than those 369 businesses receiving a "Conditional Pass." "Conditional pass" indicates that certain businesses must address specific issues to fully comply with health standards. This visual demonstrates that most businesses inspected adhere to the required health and safety guidelines.

@@

Significantly fewer infractions were labelled major than minor ones as [@tbl-total_count_severity_levels] shown, which may indicate that more serious health risks are less likely to occur. Nonetheless, the presence of these infractions indicates the need for improvements to ensure public health.

@@

```
\break
## Actions from Infraction Companies
To further evaluate the actions these violating companies took after being flagged, the report used the "Minor" level with the most violations for analysis and created a bar chart to more clearly visualize the data in [@fig-distribution_minor_severity_infractions]. Please note that the coding of this chart refers to the "Top 50 Ggplot2 Visualization" [@rstats_top50ggplot2] website. The chart is dominated by the green proportion, representing the behaviour of "notice to comply." This suggests that in most cases of minor level, immediate closure or fines are not the first action Toronto Public Health considers; instead, businesses are allowed to make amends within a certain period.
